Entrainment of low Mach number thermals in stratified domains

### 2019-05-24 tex/stratified_thermals.tex
draft 2

\subsection{Evolution of momentum and impulse}
The evolution of thermals as buoyant vortex rings has been well described in the unstratified, Boussinesq limit for decades \citep[see e.g.][for a description and sources]{lecoanet&jeevanjee2018}.
Here we lay out a description of the momentum and impulse of the thermal which will later be used to describe thermal properties like radii and depth vs. time.

In this work, we study an ideal gas and focus on the ideal, low Mach number regime in an adiabatically stratified atmosphere. 
In this regime, a linearized equation of state describes the thermodynamics well, and the fully compressible Euler momentum equation takes the form \citep{brown&all2012},
\begin{equation}
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} = 
-\grad\varpi - \frac{S_1}{c_P}\bm{g},
\label{eqn:euler_momentum}
\end{equation}
where $\bm{u}$ is the velocity, $\varpi = P_1 / \rho_0$ is the reduced pressure, $S = c_V\ln T - R\ln\rho$ is the specific entropy, and thermodynamics are broken down into background (subscript 0) and fluctuating (subscript 1) components.
In this work we find it instructive to examine the full momentum, and so multiplying this equation by the density, we obtain,
\begin{equation}
\frac{\partial (\rho\bm{u})}{\partial t} + \bm{u}\cdot\grad(\rho\bm{u}) + \rho\bm{u}(\DivU)
= -\rho\grad\varpi - \rho\frac{S_1}{c_P}\bm{g}.
\label{eqn:rho_euler_momentum}
\end{equation}
Hereafter we will define the total derivative, $D/Dt \equiv \partial/\partial t + \bm{u}\cdot\grad$, and we acknowledge that the Langrangian derivative commutes with a volume integral such that
\begin{equation*}
\frac{d}{dt}\int_{\mathcal{V}} f dV = \int_{\mathcal{V}} \left[\frac{Df}{Dt} + f(\DivU)\right]dV,
\end{equation*}
assuming that the volume $\mathcal{V}$ is advected with the fluid.
Volume-integrating Eqn.~\ref{eqn:rho_euler_momentum}, we thus find
\begin{equation}
\frac{d\bm{M}}{dt} = \int_{\mathcal{V}}\left(-\rho\grad\varpi - \rho\frac{S_1}{c_P}\bm{g}\right)dV,
\label{eqn:int_momentum_eqn}
\end{equation}
where the volume-integrated momentum is defined $\bm{M} \equiv \int_{\mathcal{V}}\rho\bm{u} dV$.
At this point we will make the assumption of a plane-parallel atmosphere in which the gravity is constant, $\bm{g} = -g\hat{z}$, and note that the z-component of the volume-integrated momentum evolves according to
\begin{equation}
\frac{d M_z}{dt} = \int_{\mathcal{V}}\left( -\rho\frac{\partial \varpi}{\partial z} + \rho g \frac{S_1}{c_P} \right)dV.
\label{eqn:Mz_definition}
\end{equation}
At this point, we find it useful to define the total buoyancy,
\begin{equation}
B \equiv \int_{\mathcal{V}} \rho\, S_1\, \frac{g}{c_P}\, dV,
\label{eqn:tot_buoyancy}
\end{equation}
as in the absence of viscosity and detrainment, this is a conserved quantity during thermal evolution (REFERENCE FIGURE). 
Furthermore, in the Boussinesq limit, the work of \citet{tarshish&all2018} shows that pressure terms reduce the efficacy of the buoyancy in changing the momentum by effectively reducing the buoyant acceleration by some factor $\beta \sim 0.5$. 
As a result, we can approximate the growth of a thermal's momentum according to
\begin{equation}
\frac{d M_z}{dt} \approx \beta B.
\label{eqn:theory_momentum}
\end{equation}

After describing the growth of the thermal's integrated momentum with time, we turn our attention to the hydrodynamic impulse, which in the stratified limit is defined
\citep{shivamoggi2010},
\begin{equation}
\bm{I} = \frac{1}{2}\int_{\mathcal{V}} \bm{x}\times(\grad\times(\rho\bm{u}))dV,
\end{equation}
where $\bm{x}$ is the position vector. 
Impulse is equivalent to the time-integrated work acting on the fluid resulting in the current fluid motion. 
Per \citet{shivamoggi2010}, changes in the impulse can be expressed as
\begin{equation*}
\frac{d\bm{I}}{d t} = \int_{\mathcal{V}}\frac{\partial(\rho\bm{u})}{\partial t}dV
= B\hat{z} - \int_{\mathcal{V}}\left[\rho\grad\varpi + \Div{\rho\bm{u}\bm{u}}\right]dV ,
\end{equation*}
where, importantly, the \emph{eulerian} time derivative of the momentum is inside of the volume integral here. 
Under the proper specification of boundary conditions, the integral term is zero, and the vertical impulse of a thermal thus straightforwardly changes in time as
\begin{equation}
\frac{d I_z}{d t} = B.
\label{eqn:change_in_impulse}
\end{equation}
We have thus retrieved the two findings on which our theory of thermal evolution will be built: \emph{both the impulse and momentum experience constant changes in time determined by the buoyant nature of the thermal}.

In the low-Mach number limit in which changes in density from the background are negligible, and in the limit of a thin-core vortex ring, the impulse of a vortex ring can be approximated as
\begin{equation}
I_z \approx \pi \rho r^2 \Gamma,
\label{eqn:impulse_approx}
\end{equation}
where $r$ is the radius of the thermal from its axis of symmetry to the maxima of its buoyant signature, and $\Gamma = \int_{\mathcal{A}} (\grad\times\bm{u})dA$ is the integrated circulation in a cross-section of the vortex ring.
Through integration of the momentum equation, it can be shown that in the absence of viscous effects, changes in circulation in an axisymmetric vortex ring are
\begin{equation}
\frac{d\Gamma}{dt} = \oint_{\mathcal{C}} g \frac{S_1}{c_P}\hat{z} \cdot d\bm{x},
\label{eqn:circulation}
\end{equation}
where $\mathcal{C}$ is the contour around the area $\mathcal{A}$ in which the circulation is contained and $d\bm{x}$ is the line integral element around that contour.
For the case of a vortex core in which the entropic signature is contained tightly in the core, as in Fig.~\ref{fig:evolution_colormeshes}, any contour which is drawn around the vortex core should conserve circulation in the absence of viscosity.

\subsection{Parameterized description of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.png}
    \caption{
	Plotted are time traces for three quantities which we assume to be constant in our thermal theory. 
	In (a) and (b), the circulation and buoyancy, divided by their maxima, are plotted vs. depth. 
	We see that with increasing stratification, there is marginally more detrainment of both of these quantities, but to first order they are constant over the	evolution of the thermal. 
	In (c) the constant $f = V/r^3$ is plotted; after significant noise during the development of the vortex ring, this quantity remains relatively constant, and increasingly approaches a constant with depth.
    \label{fig:constants} }
\end{figure}

After spinning up, we assume that the thermal is characterized by a constant buoyant signature, $B_0$, and constant circulation, $\Gamma_0$.
While these thermals began as initial spherical perturbations, they can be modeled as vortex rings which evolved from a ``virtual origin'' at which they had no radius. 
From this virtual origin, the vortex ring's momentum and impulse grow to a value of $M_0$ and $I_0$ at time $t = 0$ when the true thermal is released from rest.

We first note, as in Fig.~\ref{fig:constants}a, that the buoyancy in our thermals is not necessarily perfectly constant in time, and that effects like detrainment result in a minor reduction of the buoyant signature of our thermals. 
We thus express the buoyancy as 
\begin{equation}
B \approx \chi B_0,
\end{equation}
where $\chi$ is a constant of O(1) which represents this detrainment. 
We then integrate Eqn.~\ref{eqn:change_in_impulse} in time,
\begin{equation*}
I_z = \chi B_0 t + I_0.
\label{eqn:impulse_line}
\end{equation*}

Following Eqn.~\ref{eqn:circulation}, we take $\Gamma_0$ to be constant, which is a decent assumption in our numerical experiments, see Fig.~\ref{fig:constants}b.
Combining Eqns.~\ref{eqn:impulse_line} and \ref{eqn:impulse_approx}, we retrieve our first result,
\begin{equation}
r = \sqrt{\frac{\chi B_0 t + I_0}{\pi\rho\Gamma_0}}.
\label{eqn:r_theory}
\end{equation}
In the Boussinesq limit where $\rho \rightarrow \text{constant}$, we retrieve the $r \propto \sqrt{t}$ scaling found in the Boussinesq regime by \citet{lecoanet&jeevanjee2018}. 
We find that the inclusion of stratification adds the additional complexity of $r \propto \rho^{-1/2}$, such that downward-propagating vortex rings (as studied here) will entrain less than boussinesq thermals, and upward-propagating rings will entrain more. 
In the absence of buoyancy ($B = 0$), this result aligns with the prediction for purely horizontal compression noted by \citet{brandenburg2016} of $r^2 \propto \rho^{-1}$.

The momentum can likewise be integrated like the impulse,
\begin{equation*}
M_z = \beta\chi B_0 t + M_0.
\end{equation*}
For the proper choice of vertical velocity, $w_{\text{th}}$, and volume, $\mathcal{V}$, the momentum can be expressed precisely as 
\begin{equation*}
M_z = \rho \mathcal{V} w_{\text{th}},
\end{equation*}
and the volume can be approximated as $\mathcal{V} = f r^3$, where $f$ is a parameter which we take to be constant. 
This assumption is perhaps not as perfect as the assumptions of constant $B_0$ or $\Gamma_0$ (see Fig.~\ref{fig:constants}c), but it is nevertheless not a bad assumption. 
Combining our approximate expressions, and inserting our theoretical description of $r$ (Eqn.~\ref{eqn:r_theory}), we retrieve
\begin{equation}
\rho^{-1/2} w_{\text{th}} = \left(\frac{(\pi \Gamma)^{3/2}}{f}\right)\frac{\beta\chi B_0 t + M_0}{(\chi B_0 t + I_0)^{3/2}}.
\end{equation}
Defining the thermal velocity $w_{\text{th}} \equiv dz_{\text{th}}/dt$, and making the assumption that the vortex ring starts at a ``virtual origin'' (CITE) at $t = 0$ where it has no impulse or momentum ($I_0 = M_0 = 0$), an integrable expression for the evolution of thermal position over time can be retrieved,
\begin{equation}
\frac{dz_{\text{th}}}{\rho(z_{\text{th}})^{1/2}} =
\left(\frac{\beta(\pi\Gamma)^{3/2}}{f(\chi B_0)^{1/2}}\right)\frac{dt}{t^{1/2}}
\label{eqn:dz_theory}
\end{equation}
If the atmospheric stratification in which the thermal is falling is known, this result can be integrated with $\rho(z_{\text{th}})$ plugged in in order to find the position of the thermal as a function of time. 
We leave this result general for now, and will integrate it using the polytropic stratification used in our simulations at the end of section \ref{sec:experiment}.

\subsection{Solution for thermal evolution in a Polytrope}
Throughout this work we will study a polytrope, as is specified later in Eqn.~\ref{eqn:polytrope}.
Integrating Eqn.~\ref{eqn:dz_theory} with this specific density stratification, we find
\begin{equation}
z_{\text{th}} = \grad_{\text{ad}}^{-1}\left[\left(\frac{2C}{ \alpha } \sqrt{t + t_{\text{off}}} + T_0^{1/\alpha}  \right)^{\alpha} - 1\right] + L_z,
\label{eqn:theory_z}
\end{equation}
where $C \equiv \beta \pi^{3/2} \grad_{\text{ad}} / f \sqrt{\Gamma^3/(\chi B)}$, the temperature at the virtual origin is $T_0 = 1 + \grad_{\text{ad}}(z_0 - L_z)$, $\alpha^{-1} = 1 - m_{\text{ad}}/2$, and assuming that the polytropic index is $m_{\text{ad}} < 2$ (which is valid for our case of $m_{\text{ad}} = 1.5$ studied here).
In the limit of large stratification, we thus find that $z_{\text{th}} \propto t^2$ for our case of $\alpha = 4$.

The thermal is initialized as a uniform sphere of dense fluid, and it quickly spins up into a vortex ring. 
While we do not attempt to model the spin-up phase in this paper, it can be parameterized by the buoyancy $B_0$, circulation $\Gamma$, as well as the virtual origin $z_0$, and temporal offset $t_off$ which are the height and time a which the vortex ring would have had zero radius in the absence of a spin-up phase. 
Our theory also involves the volumetric aspect ratio of the thermal $f$, the detrainment fraction $\chi$, and the effective buoyancy $\beta$. 
These appear to be constant or only weakly dependent on the stratification for the thermals we have simulated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% EXPERIMENT SECTION
\section{Numerical Experiment} 
\label{sec:experiment}

\subsection{Anelastic Simulations}
In this work, we primarily study the evolution of 2D, azimuthally symmetric, anelastic thermals in cylindrical coordinates. 
We later verify that select simulations produce the same results as 3D fully compressible simulations in cartesian domains (see sec. REF). 
The LBR anelastic equations are \citep{lecoanet&all2014},
\begin{gather}
\DivU = -w \partial_z \ln\rho_0, \\
\begin{split}
\partial_t \bm{u} + \bm{u}\cdot\grad&\bm{u} = \\
- \grad \varpi + S_1\hat{z} &
+ \frac{1}{\rho_0\text{Re}}\left[\grad^2 \bm{u} + \frac{1}{3}\grad(\DivU)\right] 
\end{split}\\
\begin{split}
\partial_t S_1 + \bm{u}\cdot\grad S_1 =& \\
\frac{1}{\text{Re}}\left(\frac{1}{\text{Pr}\rho_0c_P }\right.&[\grad^2 S_1 + \partial_z\ln T_0 \cdot\partial_z S_1]\\
&+ \left.\frac{-(\grad_{\text{ad}})}{\rho_0 T_0}\sigma_{ij}\partial_{x_i}u_j \right),
\end{split}
\end{gather}
where $\grad_{\text{ad}} \equiv g(e^{n_\rho/m_{\text{ad}}} - 1)/(L_z c_P)$ and $\lilstressT$ is the viscous stress tensor in units of inverse time.
In our azimuthally symetric domain, we assume that $\partial_\phi = 0$; as the initial conditions of our simulations are at rest and have no azimuthal velocity, $u_\phi$, we explicitly impose that $u_\phi = 0$; therefore $\bm{u} = u_r \hat{r} + w\hat{z}$.

These equations have been nondimensionalized in the same manner as in \citet{lecoanet&jeevanjee2018} such that the length scale is the diameter of the initial thermal perturbation and the velocity scale is the freefall velocity. 
The timescale is thus the freefall crossing time of this unit length. 
These atmospheres are fully specific in terms of the Reynolds number and Prandtl number,
\begin{equation}
\text{Re} = \frac{u_{th} L_{th}}{\nu}, \qquad
\text{Pr} = \frac{u_{th} L_{th}}{\chi}, \qquad
u_{th}^2 = \frac{g L_{th} \Delta s}{c_P},
\end{equation}
where $u_{th}$ is the freefall velocity, $L_{th}$ is the thermal length scale, and
$\Delta s$ is the magnitude of the specific entropy signature of the thermal.

@@

\subsection{Atmosphere \& Initial conditions}
We study an ideal gas whose equation of state is $P = \rho T$ and whose stratification is a perfectly adiabatic polytrope,
\begin{gather}
T_0 = 1 + (\grad_{\text{ad}})(z - L_z) \\
\rho_0 = T_0^{\,m_{\text{ad}}},
\label{eqn:polytrope}
\end{gather}
where $m_{\text{ad}} = (\gamma-1)^{-1}$, and the adiabatic temperature gradient in these nondimensional atmospheres is set with $g = m_{\text{ad}} + 1$ and $\tilde{L}_{\text{th}} = (e^{n_\rho/m_{\text{ad}}} - 1)/L_z$, where $n_\rho$ is the number of density scale heights spanned by the atmosphere and $L_z = 20$ is the nondimensional depth of the atmosphere in units of thermal diameters.

To initialize the simulation, we specify a spherical initial specific entropy perturbation,
\begin{equation}
S_1 = - \frac{A}{2}\left[1 - \text{erf}\left(\frac{r' - r_{th}}{\delta}\right)\right],
\label{eqn:thermal_IC}
\end{equation}
where $A = 1$ for our scaled equations.
Here, $r' = \sqrt{r^2 + (z - z_0)^2}$, where $z_0 = L_z - 3r_{\text{th}}$, with the thermal radius set as $r_{th} = 0.5$, and a smoothing width, $\delta = 0.1$.

@@

In setting the specific entropy to an equivalent condition to that specified in Eqn.~\ref{eqn:thermal_IC}, we note that it is essential that the initial perturbation be in pressure equilibrium. 
The set of initial conditions that achieves this is
\begin{equation}
\ln\rho_1 = S_1/c_P, \qquad T_1 = T_0(e^{-\ln\rho_1} - 1).
\end{equation}

\subsection{Numerics}
We evolve our simulations forward in time using the  Dedalus\footnote{\url{http://dedalus-project.org/}} pseudospectral framework \citep{burns&all2016} to time-evolve our equations. 
For our 2D simulations, we use using an implicit-explicit (IMEX), third-order, four-stage Runge-Kutta timestepping scheme RK443 \citep{ascher&all1997}, and for our 3D simulations we use the second order semi-implicit backward differentiation formulation SBDF2 \citep{wang&ruuth2008}.

>>>>

\subsection{Evolution of momentum and impulse}
The evolution of thermals as buoyant vortex rings has been well described in the unstratified, Boussinesq limit for decades (see e.g. \LJ\,for a description and references).
While many descriptions of vortex rings assume that they evolve in a self-similar state, we will lay out a description of the momentum and impulse of the thermal and show how these quantities control the evolution of the vortex ring.

We study a fluid characterized by an ideal gas equation of state and focus on the ideal, low Mach number regime in an adiabatically stratified atmosphere. 
In this regime, a linearized equation of state describes the thermodynamics well, and the fully compressible Euler momentum equation takes the form \citep{brown&all2012},
\begin{equation}
\frac{\partial \bm{u}}{\partial t} + \bm{u}\cdot\grad\bm{u} = 
-\grad\varpi - \frac{S_1}{c_P}\bm{g},
\label{eqn:euler_momentum}
\end{equation}
where $\bm{u}$ is the velocity, $\varpi = P_1 / \rho_0$ is the reduced pressure, $S = c_V\ln T - R\ln\rho$ is the specific entropy, and thermodynamics are broken down into background (subscript 0) and fluctuating (subscript 1) components.
In this limit where the linearized equation of state is valid, $S_1 \approx c_V T_1/T_0 - R \rho_1/\rho_0$.
In this work we find it instructive to examine the full momentum, and so multiplying this equation by the full density, we obtain,
\begin{equation}
\frac{\partial (\rho\bm{u})}{\partial t} + \bm{u}\cdot\grad(\rho\bm{u}) + \rho\bm{u}(\DivU)
= -\rho\grad\varpi - \rho\frac{S_1}{c_P}\bm{g}.
\label{eqn:rho_euler_momentum}
\end{equation}
Hereafter we will define the total derivative, $D/Dt \equiv \partial/\partial t + \bm{u}\cdot\grad$, and we acknowledge that the Langrangian derivative commutes with a volume integral according to 
\begin{equation*}
\frac{d}{dt}\int_{\mathcal{V}} f dV = \int_{\mathcal{V}} \left[\frac{Df}{Dt} + f(\DivU)\right]dV,
\end{equation*}
for a volume, $\mathcal{V}$, which is advected with the fluid.
Volume-integrating Eqn.~\ref{eqn:rho_euler_momentum}, we thus find
\begin{equation}
\frac{d\bm{M}}{dt} = \int_{\mathcal{V}}\left(-\rho\grad\varpi - \rho\frac{S_1}{c_P}\bm{g}\right)dV,
\label{eqn:int_momentum_eqn}
\end{equation}
where the volume-integrated momentum is defined $\bm{M} \equiv \int_{\mathcal{V}}\rho\bm{u} dV$.
At this point we will make the assumption of a plane-parallel atmosphere in which the gravity is constant, $\bm{g} = -g\hat{z}$, and note that the z-component of the volume-integrated momentum evolves according to
\begin{equation}
\frac{d M_z}{dt} = \int_{\mathcal{V}}\left( -\rho\frac{\partial \varpi}{\partial z} + \rho g \frac{S_1}{c_P} \right)dV.
\label{eqn:Mz_definition}
\end{equation}
At this stage, it is useful to define the total buoyancy,
\begin{equation}
B \equiv \int_{\mathcal{V}} \rho\, S_1\, \frac{g}{c_P}\, dV,
\label{eqn:tot_buoyancy}
\end{equation}
as in the absence of viscosity and detrainment, this is a conserved quantity during thermal evolution (see e.g., Fig.~\ref{fig:constants}a). 
Furthermore, in the Boussinesq limit, the work of \citet{tarshish&all2018} shows that pressure terms reduce the efficacy of the buoyancy in changing the momentum by effectively reducing the buoyant acceleration by some factor $\beta \sim 0.5$. 
As a result, we approximate the growth of a thermal's vertical momentum according to
\begin{equation}
\frac{d M_z}{dt} \approx \beta B.
\label{eqn:theory_momentum}
\end{equation}

We now turn our attention to the hydrodynamic impulse, which in the stratified limit can be expressed as \citep{shivamoggi2010},
\begin{equation}
\bm{I} = \frac{1}{2}\int_{\mathcal{V}} \bm{x}\times(\grad\times(\rho\bm{u}))dV,
\end{equation}
where $\bm{x}$ is the position vector. 
The hydrodynamic impulse is the time-integrated work which has acted on the fluid to result in the current fluid motion. 
Per \citet{shivamoggi2010}, changes in the impulse can be expressed as
\begin{equation*}
\frac{d\bm{I}}{d t} = \int_{\mathcal{V}}\frac{\partial(\rho\bm{u})}{\partial t}dV
= B\hat{z} - \int_{\mathcal{V}}\left[\rho\grad\varpi + \Div{\rho\bm{u}\bm{u}}\right]dV ,
\end{equation*}
where, importantly, the \emph{eulerian} time derivative of the momentum is inside of the volume integral here. 
Under the proper specification of volumetric boundaries, the integral term is zero, and the vertical impulse of a thermal thus straightforwardly changes in time as
\begin{equation}
\frac{d I_z}{d t} = B.
\label{eqn:change_in_impulse}
\end{equation}
We have thus retrieved the two findings on which our theory of thermal evolution will be built: \emph{both the impulse and momentum experience constant changes in time determined by the buoyant nature of the thermal}.

In the low-Mach number limit in which changes in density from the background are negligible, and in the limit of a thin-core vortex ring, the impulse of a vortex ring can be approximated as
\begin{equation}
I_z \approx \pi \rho r^2 \Gamma,
\label{eqn:impulse_approx}
\end{equation}
where $r$ is the radius of the thermal from its axis of symmetry to the maxima of its buoyant signature, and $\Gamma = \int_{\mathcal{A}} (\grad\times\bm{u})dA$ is the circulation in a cross-section of the vortex ring.
Integrating the momentum equation in the absence of viscous effects shows that changes in circulation in an axisymmetric vortex ring are
\begin{equation}
\frac{d\Gamma}{dt} = \oint_{\mathcal{C}} g \frac{S_1}{c_P}\hat{z} \cdot d\bm{x},
\label{eqn:circulation}
\end{equation}
where $\mathcal{C}$ is the contour around the area $\mathcal{A}$ in which the circulation is contained and $d\bm{x}$ is the line integral element around that contour.
For the case of a vortex core in which the entropic signature is contained tightly in the core, as in Fig.~\ref{fig:evolution_colormeshes}, any contour which is drawn around the vortex core should conserve circulation in the absence of viscosity.
As a result, we will treat the circulation of a developed vortex ring as a conserved quantity.

\subsection{Parameterized description of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.png}
    \caption{
	Evolution traces are shown for  three quantities which we assume to be constant in our parameterized thermal theory. 
	The circulation (a) and buoyancy (b), divided by their parameterized value for each case (see Table \ref{table:parameters}), are plotted vs. depth. 
	We see that with increasing stratification, there is marginally more detrainment of both of these quantities, but to first order they are constant over the	thermal's transit time. 
	In (c) the volumetric constant $f$ is shown; after significant noise during the development of the vortex ring, this quantity settles toward a value of O(7) throughout the thermal's transit.
    \label{fig:constants} }
\end{figure}

After spinning up, we assume that the thermal is characterized by a constant buoyant signature, $B_0$, and constant circulation, $\Gamma_0$.
While these thermals began as initial spherical perturbations, they can be modeled as having always been vortex rings which evolved from a ``virtual origin'' where the vortex ring is characterized by having a radius of zero. 
From this virtual origin, the vortex ring's momentum and impulse grow to a value of $M_0$ and $I_0$ at time $t = 0$ when the true thermal is released from rest.

We first note, as in Fig.~\ref{fig:constants}a, that the buoyancy in our thermals is not necessarily perfectly constant in time, and that effects like detrainment result in a minor reduction of the buoyant signature of our thermals. 
We thus express the buoyancy as 
\begin{equation}
B \approx \chi B_0,
\end{equation}
where $\chi$ is a constant of O(1) which represents this detrainment. 
We then integrate Eqn.~\ref{eqn:change_in_impulse} in time,
\begin{equation*}
I_z = \chi B_0 t + I_0.
\end{equation*}

Combining this result with Eqn.~\ref{eqn:impulse_approx}, we retrieve our first result,
\begin{equation}
r = \sqrt{\frac{\chi B_0 t + I_0}{\pi\rho\Gamma_0}}.
\label{eqn:r_theory}
\end{equation}
Following Eqn.~\ref{eqn:circulation}, and taking $\Gamma_0$ to be constant (which is a decent assumption in our numerical experiments, as in Fig.~\ref{fig:constants}b), we find that the radius grows with time but this growth is damped by increasing atmospheric density.
In the Boussinesq limit where $\rho \rightarrow \text{constant}$, we retrieve the $r \propto \sqrt{t}$ scaling found in the Boussinesq regime in \LJ. 
We find that the inclusion of stratification adds the additional complexity of $r \propto \rho^{-1/2}$, such that downward-propagating vortex rings (as studied here) will entrain less than boussinesq thermals, and upward-propagating rings will entrain more. 
In the absence of buoyancy ($B = 0$), this result aligns with the prediction for purely horizontal compression noted by \citet{brandenburg2016} of $r^2 \propto \rho^{-1}$.

The momentum can likewise be simply integrated,
\begin{equation*}
M_z = \beta\chi B_0 t + M_0.
\end{equation*}
For the proper choice of vertical velocity, $w_{\text{th}}$, and volume, $\mathcal{V}$, the momentum is also precisely 
\begin{equation*}
M_z = \rho \mathcal{V} w_{\text{th}}.
\end{equation*}
We approximate the volume as $\mathcal{V} = f r^3$, where $f$ is a parameter which we take to be constant. 
This assumption is perhaps not as perfect as the assumptions of constant $B_0$ or $\Gamma_0$ (see Fig.~\ref{fig:constants}c), but it is nevertheless not a bad assumption. 
Combining our approximate expressions, and using our theoretical description of $r$ (Eqn.~\ref{eqn:r_theory}), we retrieve
\begin{equation}
\rho^{-1/2} w_{\text{th}} = \left(\frac{(\pi \Gamma)^{3/2}}{f}\right)\frac{\beta\chi B_0 t + M_0}{(\chi B_0 t + I_0)^{3/2}}.
\end{equation}
Defining the thermal velocity $w_{\text{th}} \equiv dz_{\text{th}}/dt$, and making the assumption that the vortex ring is at the virtual origin at $t = 0$ so that $I_0 = M_0 = 0$, an integrable expression for the evolution of thermal position over time can be retrieved,
\begin{equation}
\frac{dz_{\text{th}}}{\rho(z_{\text{th}})^{1/2}} =
\left(\frac{\beta(\pi\Gamma)^{3/2}}{f(\chi B_0)^{1/2}}\right)\frac{dt}{t^{1/2}}
\label{eqn:dz_theory}
\end{equation}
If the atmospheric stratification in which the thermal is falling is known, this result can be integrated with the known profile of $\rho(z_{\text{th}})$ in order to find the position of the thermal as a function of time. 
We leave this result general for now, and will integrate it using the polytropic stratification used in our simulations in the next section.

\subsection{Solution for thermal evolution in a Polytrope}
In this work, we study an ideal gas whose equation of state is $P = \rho T$ and whose stratification is a perfectly adiabatic polytrope,
\begin{gather}
T_0 = 1 + (\grad_{\text{ad}})(z - L_z), \\
\rho_0 = T_0^{\,m_{\text{ad}}},
\label{eqn:polytrope}
\end{gather}
where $m_{\text{ad}} = (\gamma-1)^{-1}$, and where the adiabatic temperature gradient in our nondimensional systems (see section \ref{sec:experiment}) is $\grad_{\text{ad}} \equiv g(e^{n_\rho/m_{\text{ad}}} - 1)/(L_z c_P)$, with $g = m_{\text{ad}} + 1$ and $L_z = 20$ is the nondimensional depth of the atmosphere in units of thermal diameters.

Integrating Eqn.~\ref{eqn:dz_theory} under this polytropic density stratification and assuming $m_{\text{ad}} < 2$, we find
\begin{equation}
z_{\text{th}} = \grad_{\text{ad}}^{-1}\left[\left(\frac{2C}{ \alpha } \sqrt{t + t_{\text{off}}} + T_0^{1/\alpha}  \right)^{\alpha} - 1\right] + L_z,
\label{eqn:theory_z}
\end{equation}
where $C \equiv \beta \pi^{3/2} \grad_{\text{ad}} / f \sqrt{\Gamma^3/(\chi B)}$ and the temperature at the virtual origin is $T_0 = 1 + \grad_{\text{ad}}(z_0 - L_z)$.
The virtual origin is located at $z = z_0$ and $t = -t_{\text{off}}$ in simulation units, which are the height and time a which the vortex ring would have had zero radius in the absence of a spin-up phase.
The constant $\alpha^{-1} = 1 - m_{\text{ad}}/2$, and in the limit of large stratification, we find that $z_{\text{th}} \propto t^2$ for our case of $\alpha = 4$.

The thermal is initialized as a uniform sphere of dense fluid, and it quickly spins up into a vortex ring. 
While we do not attempt to model the spin-up phase in this paper, it can be parameterized by the buoyancy $B_0$, circulation $\Gamma_0$, as well as the virtual origin $z_0$, and temporal offset $t_{\text{off}}$. 
Our theory also involves the volumetric aspect ratio of the thermal $f$, the detrainment fraction $\chi$, and the effective buoyancy $\beta$. 
These appear to be constant or only weakly dependent on the stratification for the thermals we have simulated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% EXPERIMENT SECTION
\section{Numerical Experiment} 
\label{sec:experiment}
%In this work, we primarily study the evolution of 2D, azimuthally symmetric, anelastic thermals in cylindrical coordinates. 
As a result of our theory being largely derived in the anelastic, axisymmetric limit, we have chosen to focus the bulk of our attention on our 2D, axisymmetric anelastic simulations.
We choose to focus on 2D simulations, as they allow us to more easily study high density stratifications using highly reproducible simulations whose computational expense is low.
We verify that select anelastic simulations produce the same results as 3D fully compressible simulations in cartesian domains.

While natural processes are very turbulent, we have chosen to study the evolution of laminar thermals here. 
As we are not aware of a developed laminar theory in the stratified regime in the literature, like the one developed here, it is useful to restrict our study to the laminar case to determine the validity of our theory.
Furthermore, \LJ\, showed in the Boussinesq limit that the measured entrainment of turbulent thermal vortex rings is well described by laminar theory, and we have no reason to believe that result should not hold for the stratified case.
As a result, we leave studies of turbulent thermals to future work.

\subsection{Anelastic Simulations}
The LBR anelastic equations are \citep{lecoanet&all2014},
\begin{gather}
\DivU = -w \partial_z \ln\rho_0, \\
\begin{split}
\partial_t \bm{u} + \bm{u}\cdot\grad&\bm{u} = \\
- \grad \varpi + S_1\hat{z} &
+ \frac{1}{\rho_0\text{Re}}\left[\grad^2 \bm{u} + \frac{1}{3}\grad(\DivU)\right] 
\end{split}\\
\begin{split}
\partial_t S_1 + \bm{u}\cdot\grad S_1 =& \\
\frac{1}{\text{Re}}\left(\frac{1}{\text{Pr}\rho_0c_P }\right.&[\grad^2 S_1 + \partial_z\ln T_0 \cdot\partial_z S_1]\\
&+ \left.\frac{-(\grad_{\text{ad}})}{\rho_0 T_0}\sigma_{ij}\partial_{x_i}u_j \right),
\end{split}
\end{gather}
where $\lilstressT$ is the viscous stress tensor in units of inverse time.
In our azimuthally symetric domain, we assume that $\partial_\phi = 0$; as the initial conditions of our simulations are at rest and have no azimuthal velocity, $u_\phi$, we explicitly impose that $u_\phi = 0$; therefore $\bm{u} = u_r \hat{r} + w\hat{z}$.

These equations have been nondimensionalized in the same manner as in \citet{lecoanet&jeevanjee2018} such that the length scale is the diameter of the initial thermal perturbation and the velocity scale is the freefall velocity. 
The timescale is thus the freefall crossing time of this unit length. 
These equations are then fully specified in terms of the Reynolds number and Prandtl number,
\begin{equation}
\text{Re} = \frac{ u_{\text{th}} L_{\text{th}}}{\nu}, \qquad
\text{Pr} = \frac{ u_{\text{th}} L_{\text{th}}}{\chi}, \qquad
u_{\text{th}}^2 = \frac{g L_{\text{th}} \Delta s}{c_P},
\end{equation}
where $u_{\text{th}}$ is the freefall velocity, $L_{th}$ is the thermal length scale, and
$\Delta s$ is the magnitude of the specific entropy signature of the thermal.

@@

\subsection{Initial conditions}
To initialize the simulation, we specify a spherical initial specific entropy perturbation,
\begin{equation}
S_1 = - \frac{1}{2}\left[1 - \text{erf}\left(\frac{r' - r_{\text{th}}}{\delta}\right)\right].
\label{eqn:thermal_IC}
\end{equation}
Here, $r' = \sqrt{r^2 + (z - z_0)^2}$, where $z_0 = L_z - 3r_{\text{th}}$, with the thermal radius set as $r_{\text{th}} = 0.5$, and a smoothing width, $\delta = 0.1$.
As mentioned previously, Re = 600 and Pr = 1 are specified at the thermal's initial depth, $z = z_0$.

@@

In setting the specific entropy to an equivalent condition to that specified in Eqn.~\ref{eqn:thermal_IC}, we note that it is essential that the initial perturbation be in pressure equilibrium. 
The set of initial conditions that achieves this is
\begin{equation}
\ln\rho_1 = S_1/c_P, \qquad T_1 = T_0(e^{-\epsilon\ln\rho_1} - 1)/\epsilon.
\end{equation}

\subsection{Numerics}
We evolve our simulations forward in time using the  Dedalus\footnote{\url{http://dedalus-project.org/}} pseudospectral framework \citep{burns&all2016} [CITE THE METHODS PAPER]. 
For our 2D simulations, we use using an implicit-explicit (IMEX), third-order, four-stage Runge-Kutta timestepping scheme RK443 \citep{ascher&all1997}, and for our 3D simulations we use the second-order semi-implicit backward differentiation formulation SBDF2 \citep{wang&ruuth2008}.
